Estimating the distribution of block sizes as a function of average transaction rates

### 2015-10-21 paper/paper.tex
Some minor edits plus adding citations

\documentclass{article}
\usepackage{fullpage}
\usepackage{amsmath}
\usepackage{bm, bbm}
%\usepackage{mathtools}
\usepackage[noblocks]{authblk}

@@

\section{Introduction}
The optimal maximum size of Bitcoin blocks is an area of active controversy. When the maximum size is set at a high level, mining centralization will increase due to higher resource demands, and the network becomes vulnerable to selfish mining attacks.\footnote{http://arxiv.org/abs/1311.0243} When the maximum size is too small, there is inadequate space to accommodate the volume of transactions that users demand, resulting in higher fees, wait times, and potentially lower rates of adoption, which affects security through the price of Bitcoin. The issue may be with us for some time; even assuming that current disagreements over raising the 1MB cap are resolved, if the Bitcoin network keeps growing and as computing power increases, the issue will need to be re-litigated from time to time. One proposal, BIP 100, would subject the decision to a continuous voting process by miners indefinitely into the future.

@@

% latex table generated in R 3.2.2 by xtable 1.7-4 package
% Mon Oct 12 10:44:18 2015
\begin{table}[ht]
\centering
\begin{tabular}{lrrr}
  \hline
 & Mean & 2.5\% & 97.5\% \\ 
  \hline
       $n=5000$ & 1.34 & 1.29 & 1.39 \\ 
$n=1000$, $g=1$ & 1.26 & 1.17 & 1.37 \\ 
$n=1000$, $g=2$ & 1.36 & 1.26 & 1.47 \\ 
$n=1000$, $g=3$ & 1.37 & 1.26 & 1.47 \\ 
$n=1000$, $g=4$ & 1.27 & 1.17 & 1.37 \\ 
$n=1000$, $g=5$ & 1.41 & 1.30 & 1.52 \\ 
 $n=500$, $g=1$ & 1.29 & 1.15 & 1.45 \\ 
 $n=500$, $g=2$ & 1.23 & 1.10 & 1.37 \\ 
 $n=500$, $g=3$ & 1.46 & 1.31 & 1.63 \\ 
 $n=500$, $g=4$ & 1.28 & 1.14 & 1.42 \\ 
 $n=500$, $g=5$ & 1.55 & 1.38 & 1.74 \\ 
 $n=500$, $g=6$ & 1.21 & 1.08 & 1.34 \\ 
 $n=500$, $g=7$ & 1.24 & 1.11 & 1.38 \\ 
 $n=500$, $g=8$ & 1.29 & 1.15 & 1.44 \\ 
 $n=500$, $g=9$ & 1.36 & 1.21 & 1.52 \\ 
 $n=500$, $g=10$ & 1.46 & 1.30 & 1.63 \\ 
 \hline
\end{tabular}
\caption{Posterior mean and 95\% credible intervals for $\alpha$ from fitting a gamma distribution to block sizes of the last 5000 blocks in our database. For sample sizes less than 5000, blocks were broken into groups of equal size and the model was fit separately to each group. Sample size is indicated by $n$, group id by $g$. Each model was fit in RStan\protect\footnotemark using 4 chains, each with 4000 iterations, 2000 of which used for warmup/burn in.}
\label{tab:gammafit}
\end{table}

For\footnotetext{Cite RStan} the full 5000 blocks, the posterior mean of $\alpha$ is 1.34 and the lower bound of the 95\% credible interval is 1.29, so we have strong evidence that $\alpha$ is nonzero and that a constant transaction rate is not appropriate. This is not surprising since Bitcoin usage has gone up over time, but on a short enough time scale usage may be approximately constant. A block is around 10 minutes on average, so 5000 blocks takes a little over a month on average. It is possible that on a short enough time-scale, constant transaction rates are more appropriate. So we fit the gamma distribution to 1000 blocks at a time --- representing about a week --- by breaking the original 5000 blocks into five groups. For each of these groups the 95\% credible interval does not contain one --- the lowest lower bound is 1.17. Similarly for groups of 500 blocks, each representing about 3.5 days, the lowest lower bound is about 1.08. So we can safely conclude that the transaction rate is not constant. A likely reason for this is seasonality in transaction rates --- Bitcoin trades are probably more likely to happen during certain times of day than others, based on when relevant markets are open and when the bulk of Bitcoin users are awake. The {\it mean} transaction rate averaging over e.g. seasonality is probably approximately constant for time-scales this small, however. In any case, transaction rates are nonconstant even on short time-scales, which motivates modeling them as such.

@@

The second major constraint that the block chain imposes on us is that transaction data is only measured once a block arrives. So at random measurement times which we only observe with error, we observe the amount of transaction data since the previous measurement time --- though we observe this without error. These two constraints motivate a class of models for modeling Bitcoin transaction data. Once again, let $\tau_k$ denote the actual time of the arrival of the $k$'th block, $\delta_k$ the amount of time between block $k$ and block $k-1$ so that $\tau_{k}=\tau_{k-1} + \delta_k$, and now let $t_k$ denote the observed arrival time in the blockchain. Further, let $x_\tau$ denote the amount data in the entire chain at time $\tau$ so that $x_{\tau_{k}} - x_{\tau_{k-1}}$ denotes the amount of data in block $k$. Now suppose the amount of transaction data in MBs is distributed according to some nondecreasing, infinitely divisible\footnote{Should we define?} stochastic process. Most of the interesting modeling choices come from modeling this stochastic process. Conventional Brownian motion is inappropriate since the increments should be nonnegative while geometric Brownian motion is a poor choice since it cannot start from zero. Instead, we consider the classes of non-negative infinite activity\footnote{Should we define?} L\'{e}vy processes, often called subordinators, with closed form solutions for $x_{\tau}$.\footnote{Cite the Levy process chapter here}. The stochastic process $\{x_{\tau}:\tau>0\}$ is a L\'{e}vy process if it is continous time and has independent and stationary increments with $x_0=0$. Brownian motion is the most common example. A nondecreasing example is the Gamma process where $x_{\tau + \delta} - x_{\tau} \sim G(\alpha \delta, \beta)$ for any time $\tau$ and positive increment $\delta$.

While all L\'{e}vy processes are infinitely divisible, the gamma process is particularly convenient since the distribtion of $x_{\tau+\delta} - x_\tau$ and $x_\tau$ are both known in closed form as a consequence of the fact that if $x_1\sim G(\alpha_1,\beta)$ independent of $x_2\sim G(\alpha_2,\beta)$, then $x_1+x_2 \sim G(\alpha_1 + \alpha_2, \beta)$. Often L\'{e}vy processes are defined by considering an infinitely divisible distribution for $x_1$, but the distribution of $x_{\tau}$ for any $\tau>0$ need not be from the same class as $x_1$. The inverse Gaussian process is another L\'{e}vy process which has a closed form solution for the distribution of $x_\tau$. The inverse Gaussian density is
\[
p(x) = \left[\frac{\lambda}{2\pi x^3}\right]^{1/2}\exp\left[\frac{-\lambda(x - \mu)^2}{2\mu^2x}\right]
\]
for $x>0$ with mean parameter $\mu>0$, shape parameter $\lambda>0$, and variance $\mu^3/\lambda$. Then increments of the inverse Gaussian process are inverse Gaussian distributed, viz. $x_{\tau + \delta} - x_\tau \sim IG(\mu\delta,\lambda\delta^2)$.

It is attractive that we can write down the distribution of $x_{\tau + \delta} - x_\tau$ in closed form since the observation times are observed with error. This simplifies Markov chain Monte Carlo methods that may be used to fit the model by allowing us to exactly simulate $x_{\tau_{k-1} + \delta_k}$ without resorting to approximation methods. The parameters of the Gamma or inverse Gaussian distributions can also be further modeled as stochastic processes in order to capture e.g. seasonal variation or stochastic volatility, though most reasonable models on these parameters will not allow for closed form distributions of the parameter increments. We will use the gamma process since it is widely available in statistical packages, unlike the main alternative, the inverse Gaussian.

@@

% latex table generated in R 3.2.2 by xtable 1.7-4 package
% Mon Oct 12 22:27:11 2015
\begin{table}[ht]
\centering
\begin{tabular}{rrrrrrrrr}
  \hline
 & Mean & SD & SE(Mean) & 2.5\% & 25\% & 50\% & 75\% & 97.5\% \\ 
  \hline
$\gamma$ & 0.0240 & 0.0004 & 0.0000 & 0.0232 & 0.0237 & 0.0240 & 0.0243 & 0.0248 \\ 
  $\beta$ & 15.0951 & 0.9228 & 0.0103 & 13.3725 & 14.4461 & 15.0760 & 15.6962 & 16.9591 \\ 
  $\sigma$ & 0.0024 & 0.0001 & 0.0000 & 0.0023 & 0.0023 & 0.0024 & 0.0024 & 0.0025 \\ 
   \hline
\end{tabular}
 \caption{Results of fitting the measurement error model with stochastic measurement times. We use RStan to fit the model, obtaining four chains each of 4000 iterations, 2000 of which used for warmup/burn in.\protect\footnotemark (Note: outdated numbers, updated ones coming) }
 \label{tab:modelfit}
\end{table}

@@

Given\footnotetext{Cite RStan} an approximate simulation of the posterior distribution of the model parameters, we can use Monte Carlo simulation to estimate the probability of a block filling up in a number of ways. The simplest way is to compute parameter estimates based on the posterior, e.g. posterior means or posterior medians, then estimate the probability of a block filling up by simulating the model conditional on these parameter values. In order to take into account increased Bitcoin usage, we can manually increase the mean transaction data rate parameter, $\gamma$, while holding other parameters constant. In order to take into account seasonality, we can make $\gamma$ change in a predictable, seasonal pattern. This approach directly specifies the mean transaction rate at $\gamma$, but because of the mean-variance relationship of the gamma distribution it also indirectly specifies the variance of the transaction rate at $\gamma/\hat{beta}$ where $\hat{\beta}$ is an estimate of $\beta$.

This approach has two major shortfalls: 1) it ignores parameter uncertainty, and 2) it assumes that the other parameters will remain constant as $\gamma$ increases. We can mitigate 1) while making 2) if not worse, different. Specifically, given access to an approximate simulation of the posterior of the model parameters, we can simulate the model for each draw from the posterior in order to estimate the probability of a block filling up --- this is the posterior predictive distribution.\footnote{Cite Gelman et al. text here} In order to allow for different transaction rates in this simulation, we can set $\gamma$ at a particular value while using the full posterior of the other model parameters, or even better we can shift the posterior of $\gamma$ so that the posterior mean of the shifted posterior is the desired value. This approach takes into account parameter uncertainty but assumes that our uncertainty about the other parameters would be unchanged if the location of $\gamma$'s distribution were different. Essentially this requires that if we were in a higher transaction rate regime, our priors would have been different such that the posterior distribution would have obtained would be the shifted posterior we construct. Explicitly constructing this prior would be involved and maybe intractable or even impossible. Instead we estimate the probability of a block filling using all three approaches, but acknowledge that the estimates become less credible for values of $\gamma$ or $\mathrm{E}[\gamma]$ further away from their estimated values.

@@

>>>>

\documentclass{article}
\usepackage{fullpage}
\usepackage{amsmath}
\usepackage{bm, bbm}
\usepackage[authoryear]{natbib} 
\usepackage[noblocks]{authblk}

@@

\section{Introduction}
The optimal maximum size of Bitcoin blocks is an area of active controversy. When the maximum size is set at a high level, mining centralization will increase due to higher resource demands, and the network becomes vulnerable to selfish mining attacks\citep{eyal2014majority}.%\footnote{http://arxiv.org/abs/1311.0243}
 When the maximum size is too small, there is inadequate space to accommodate the volume of transactions that users demand, resulting in higher fees, wait times, and potentially lower rates of adoption, which affects security through the price of Bitcoin. The issue may be with us for some time; even assuming that current disagreements over raising the 1MB cap are resolved, if the Bitcoin network keeps growing and as computing power increases, the issue will need to be re-litigated from time to time. One proposal, BIP 100, would subject the decision to a continuous voting process by miners indefinitely into the future.

@@

% latex table generated in R 3.2.2 by xtable 1.7-4 package
% Mon Oct 12 10:44:18 2015
\begin{table}[ht]
\centering
\begin{tabular}{lrrr}
  \hline
 & Mean & 2.5\% & 97.5\% \\ 
  \hline
       $n=5000$ & 1.34 & 1.29 & 1.39 \\ 
$n=1000$, $g=1$ & 1.26 & 1.17 & 1.37 \\ 
$n=1000$, $g=2$ & 1.36 & 1.26 & 1.47 \\ 
$n=1000$, $g=3$ & 1.37 & 1.26 & 1.47 \\ 
$n=1000$, $g=4$ & 1.27 & 1.17 & 1.37 \\ 
$n=1000$, $g=5$ & 1.41 & 1.30 & 1.52 \\ 
 $n=500$, $g=1$ & 1.29 & 1.15 & 1.45 \\ 
 $n=500$, $g=2$ & 1.23 & 1.10 & 1.37 \\ 
 $n=500$, $g=3$ & 1.46 & 1.31 & 1.63 \\ 
 $n=500$, $g=4$ & 1.28 & 1.14 & 1.42 \\ 
 $n=500$, $g=5$ & 1.55 & 1.38 & 1.74 \\ 
 $n=500$, $g=6$ & 1.21 & 1.08 & 1.34 \\ 
 $n=500$, $g=7$ & 1.24 & 1.11 & 1.38 \\ 
 $n=500$, $g=8$ & 1.29 & 1.15 & 1.44 \\ 
 $n=500$, $g=9$ & 1.36 & 1.21 & 1.52 \\ 
 $n=500$, $g=10$ & 1.46 & 1.30 & 1.63 \\ 
 \hline
\end{tabular}
\caption{Posterior mean and 95\% credible intervals for $\alpha$ from fitting a gamma distribution to block sizes of the last 5000 blocks in our database. For sample sizes less than 5000, blocks were broken into groups of equal size and the model was fit separately to each group. Sample size is indicated by $n$, group id by $g$. Each model was fit in RStan \citep{rstan-software:2015,stan-software:2015} using 4 chains, each with 4000 iterations, 2000 of which used for warmup/burn in.}
\label{tab:gammafit}
\end{table}

For the full 5000 blocks, the posterior mean of $\alpha$ is 1.34 and the lower bound of the 95\% credible interval is 1.29, so we have strong evidence that $\alpha$ is nonzero and that a constant transaction rate is not appropriate. This is not surprising since Bitcoin usage has gone up over time, but on a short enough time scale usage may be approximately constant. A block is around 10 minutes on average, so 5000 blocks takes a little over a month on average. It is possible that on a short enough time-scale, constant transaction rates are more appropriate. So we fit the gamma distribution to 1000 blocks at a time --- representing about a week --- by breaking the original 5000 blocks into five groups. For each of these groups the 95\% credible interval does not contain one --- the lowest lower bound is 1.17. Similarly for groups of 500 blocks, each representing about 3.5 days, the lowest lower bound is about 1.08. So we can safely conclude that the transaction rate is not constant. A likely reason for this is seasonality in transaction rates --- Bitcoin trades are probably more likely to happen during certain times of day than others, based on when relevant markets are open and when the bulk of Bitcoin users are awake. The {\it mean} transaction rate averaging over e.g. seasonality is probably approximately constant for time-scales this small, however. In any case, transaction rates are nonconstant even on short time-scales, which motivates modeling them as such.

@@

The second major constraint that the block chain imposes on us is that transaction data is only measured once a block arrives. So at random measurement times which we only observe with error, we observe the amount of transaction data since the previous measurement time --- though we observe this without error. These two constraints motivate a class of models for modeling Bitcoin transaction data. Once again, let $\tau_k$ denote the actual time of the arrival of the $k$'th block, $\delta_k$ the amount of time between block $k$ and block $k-1$ so that $\tau_{k}=\tau_{k-1} + \delta_k$, and now let $t_k$ denote the observed arrival time in the blockchain. Further, let $x_\tau$ denote the amount data in the entire chain at time $\tau$ so that $x_{\tau_{k}} - x_{\tau_{k-1}}$ denotes the amount of data in block $k$. Now suppose the amount of transaction data in MBs is distributed according to some nondecreasing stochastic process. Most of the interesting modeling choices come from modeling this stochastic process. Conventional Brownian motion is inappropriate since the increments should be nonnegative while geometric Brownian motion is a poor choice since it cannot start from zero. Instead, we consider non-negative infinite activity L'{e}vy processes, often called subordinators, with closed form solutions for $x_{\tau}$. The following discussion is brief -- see e.g. \citet{barndorff2012basics} for a fuller treatment. The stochastic process $\{x_{\tau}:\tau>0\}$ is a L\'{e}vy process if it is continous time and has independent and stationary increments with $x_0=0$. Brownian motion is the most common example. A nondecreasing example is the Gamma process where $x_{\tau + \delta} - x_{\tau} \sim G(\alpha \delta, \beta)$ for any time $\tau$ and positive increment $\delta$.

The gamma process is particularly convenient since the distribtion of $x_{\tau+\delta} - x_\tau$ and $x_\tau$ are both known in closed form as a consequence of the fact that if $x_1\sim G(\alpha_1,\beta)$ independent of $x_2\sim G(\alpha_2,\beta)$, then $x_1+x_2 \sim G(\alpha_1 + \alpha_2, \beta)$. Often L\'{e}vy processes are defined by considering an infinitely divisible distribution for $x_1$, but the distribution of $x_{\tau}$ for any $\tau>0$ need not be from the same class as $x_1$. The inverse Gaussian process is another L\'{e}vy process which has a closed form solution for the distribution of $x_\tau$ and for which $x_\tau$ is inverse Gaussian distributed for all $\tau>0$. The inverse Gaussian density is
\[
p(x) = \left[\frac{\lambda}{2\pi x^3}\right]^{1/2}\exp\left[\frac{-\lambda(x - \mu)^2}{2\mu^2x}\right]
\]
for $x>0$ with mean parameter $\mu>0$, shape parameter $\lambda>0$, and variance $\mu^3/\lambda$. Then increments of the inverse Gaussian process are inverse Gaussian distributed, viz. $x_{\tau + \delta} - x_\tau \sim IG(\mu\delta,\lambda\delta^2)$.

It is attractive that we can write down the distribution of $x_{\tau + \delta} - x_\tau$ in closed form since the observation times are observed with error. This simplifies Markov chain Monte Carlo methods that may be used to fit the model by allowing us to exactly simulate $x_{\tau_{k-1} + \delta_k}$ without resorting to approximation methods. The parameters of the Gamma or inverse Gaussian distributions can also be further modeled as stochastic processes in order to capture e.g. seasonal variation or stochastic volatility, though most reasonable models on these parameters will not allow for closed form distributions of the parameter increments. We will use the gamma process since it is widely available in statistical packages and in particular is available in Stan \citep{stan-software:2015}, unlike the main alternative, the inverse Gaussian.

@@

% latex table generated in R 3.2.2 by xtable 1.7-4 package
% Mon Oct 12 22:27:11 2015
\begin{table}[ht]
\centering
\begin{tabular}{rrrrrrrrr}
  \hline
 & Mean & SD & SE(Mean) & 2.5\% & 25\% & 50\% & 75\% & 97.5\% \\ 
  \hline
$\gamma$ & 0.0240 & 0.0004 & 0.0000 & 0.0232 & 0.0237 & 0.0240 & 0.0243 & 0.0248 \\ 
  $\beta$ & 15.0951 & 0.9228 & 0.0103 & 13.3725 & 14.4461 & 15.0760 & 15.6962 & 16.9591 \\ 
  $\sigma$ & 0.0024 & 0.0001 & 0.0000 & 0.0023 & 0.0023 & 0.0024 & 0.0024 & 0.0025 \\ 
   \hline
\end{tabular}
 \caption{Results of fitting the measurement error model with stochastic measurement times. We use RStan \citep{rstan-software:2015,stan-software:2015} to fit the model, obtaining four chains each of 4000 iterations, 2000 of which used for warmup/burn in.}
 \label{tab:modelfit}
\end{table}

@@

Given an approximate simulation of the posterior distribution of the model parameters, we can use Monte Carlo simulation to estimate the probability of a block filling up in a number of ways. The simplest way is to compute parameter estimates based on the posterior, e.g. posterior means or posterior medians, then estimate the probability of a block filling up by simulating the model conditional on these parameter values. In order to take into account increased Bitcoin usage, we can manually increase the mean transaction data rate parameter, $\gamma$, while holding other parameters constant. In order to take into account seasonality, we can make $\gamma$ change in a predictable, seasonal pattern. This approach directly specifies the mean transaction rate at $\gamma$, but because of the mean-variance relationship of the gamma distribution it also indirectly specifies the variance of the transaction rate at $\gamma/\hat{beta}$ where $\hat{\beta}$ is an estimate of $\beta$.

This approach has two major shortfalls: 1) it ignores parameter uncertainty, and 2) it assumes that the other parameters will remain constant as $\gamma$ increases. We can mitigate 1) while making 2) if not worse, different. Specifically, given access to an approximate simulation of the posterior of the model parameters, we can simulate the model for each draw from the posterior in order to estimate the probability of a block filling up --- this is the posterior predictive distribution \citep{gelman2014bayesian}. In order to allow for different transaction rates in this simulation, we can set $\gamma$ at a particular value while using the full posterior of the other model parameters, or even better we can shift the posterior of $\gamma$ so that the posterior mean of the shifted posterior is the desired value. This approach takes into account parameter uncertainty but assumes that our uncertainty about the other parameters would be unchanged if the location of $\gamma$'s distribution were different. Essentially this requires that if we were in a higher transaction rate regime, our priors would have been different such that the posterior distribution would have obtained would be the shifted posterior we construct. Explicitly constructing this prior would be involved and maybe intractable or even impossible. Instead we estimate the probability of a block filling using all three approaches, but acknowledge that the estimates become less credible for values of $\gamma$ or $\mathrm{E}[\gamma]$ further away from their estimated values.

@@

\bibliographystyle{apalike} 
\bibliography{btc}
